Adaptive evolution with aneuploidy and mutation

### 2021-11-17 ms/ms.tex
small grammar fixes

\paragraph{Parameter inference.} To infer model parameters, we use approximate Bayesian computation~\citep{Sunnaker2013} with a sequential Monte-Carlo scheme, or ABC-SMC~\citep{Sisson2009}, implemented in the \texttt{pyABC}\footnote{https://pyabc.readthedocs.io} Python 
package~\citep{Klinger2018}.
Briefly, this approach uses numerical stochastic simulations of the model to infer a posterior distribution over the model parameters. It is a method of likelihood-free, simulation-based inference~\citep{Cranmer2020}, that is, for estimating a posterior distribution when a likelihood function cannot be computed. It is therefore suitable to our case, in which the likelihood function can only be approximated from simulations, and cannot be directly computed. 
%adapted description from Syga et al. 2021:
In short, the algorithm employs sequential importance sampling over generations. In generation $t$, the algorithm draws sets of parameters $\theta$ from a given proposal distribution and consequently simulates data $C$ from the model until $n_t$ instances were accepted ($n_t$ "particles") based on the comparison to observed data $C_{obs}$ via a distance function $D(C, C_obs)$ and acceptance threshold $\epsilon_t$, such that $D(C, C_objs)\leq{\epsilon_t}$. As the distance function we chose $1-\likelihood$. For the acceptance threshold the median of the distances of the accepted instances of the previous generation were applied (for the first generation the acceptance threshold is defined $0.01$). 
When $n_t$ instances were accepted new proposal distribution is constructed by multivariate normal kernel with adaptive covariance matrix. In the first generation the proposal distribution equals to the prior. 
The size of $n_t$ was determined by adaptive population strategy~\citep{Klinger2017}.
We stop the iterations, when no more sufficient change in acceptance threshold occur, and no change in posterior.
After receiving weighted particles of the posterior distribution using ABC-SMC scheme, we calculate Kernel Density Estimate using Gaussian kernels. $50,000$ sampled points from KDE represent the posterior distribution over the model parameters, and we use it for plots, median calculation, etc.
% TODO convergence tests, etc. Number of replicates for different models.

@@

We used growth curves of $2n+1$ and $2n$ in \SI{39}{\celsius} to estimate  $w_{2n+1}$. We used the same prior for $w_{2n+1^*}$ and $w_{2n^*}$, assuming that fitness values shouldn't be differ too much. We also tested different priors, but this one (let's call it main prior) was better in sense of WAIC, prediction plots, estimated parameters. %TODO provide formula of WAIC. Table? To show ppc plots also?. 
\\We tested uninformative uniform prior, $\mathit{Uniform}(1,6)$, for all $w_{2n+1}$, $w_{2n+1^*}$, $w_{2n^*}$ and only for $w_{2n+1^*}$, $w_{2n^*}$, leaving $w_{2n+1}$ prior as before . In both scenarios the estimated by the model parameters of fitness were looks too high: for the first scenario the estimated medians were  $w_{2n+1}=2.767$, $w_{2n+1^*}=5.79$, $w_{2n^*}=5.827$ and for the second $w_{2n+1}=1.046$, $w_{2n+1^*}=4.642$, $w_{2n^*}=4.684$. We don't expect to have such high fitness values. Moreover the WAIC for the models with these priors are 0.57 and 1.9 respectively, while WAIC for the main prior model is 0.27. Also for the model with uninformative prior for all $w_{2n+1}$, $w_{2n+1^*}$, $w_{2n^*}$ the fixation time of 2n+1 is about 16 in average, that contradicts unpublished observations that were about 250-300.
\\Also we tried to use additional growth curves. We used growth curves of 2n* (\emph{refined} from \citet{Yona2012}) and 2n+1 in \SI{39}{\celsius} to estimate $w_{2n*}/w_{2n+1}$. The same prior was used for $w_{2n*}/w_{2n+1*}$. This model didn't perform well, resulted in WAIC 2.8.
\\Another try was to use growth curves of $2n+1$ and $2n$ in \SI{30}{\celsius} to estimate $1-c$. This estimation assumes that the cost of aneuploidy is the same in \SI{39}{\celsius} and \SI{30}{\celsius}; this might be incorrect, but we only assumed this to generate a prior distribution for the fitness values. The prior for $b$ was taken the same as for $c$. This model didn't work at all: we didn't receive any parameter that the likelihood with it is greater than zero. %TODO it make sense because bla bla?

Because \texttt{Curveball} uses a maximum-likelihood approach to estimate model parameters, we were able to estimate a distribution of relative fitness values by sampling from a truncated multivariate normal distribution defined by the maximum-likelihood covariance matrix. Thus, we sampled 10,000 values for appropriate fitness, which we used as prior distributions. 
See Figures~\ref{fig:growth-curves-30deg} and~\ref{fig:growth-curves-39deg}.

For the mutation rate $\mu$ and aneuploidy rate $\delta$ we used uniform priors, $\mathit{Uniform}(1e-8,1e-5)$ and $\mathit{Uniform}(1e-6,1e-3)$ respectively. We have tried also uniform priors with lower than $1e-8$ bound for $\mu$, but anyway the posterior ends with higher values. 
For the single locus model without aneuploidy, for the mutation rate wider uniform prior was used - $\mathit{Uniform}(1e-10,1e-5)$. 
%TODO ILIA please report here the point estimate and 95% CI for the three distributions, I only have for 1-c+b: 0.9863 (95% CI 0.9790 - 0.9939).

\paragraph{Model comparison.}  % TODO ILIA what did you do for model comparison/selection?

\pagebreak
% Results
\section*{Results}

>>>>

\paragraph{Parameter inference.} To infer model parameters, we use approximate Bayesian computation~\citep{Sunnaker2013} with a sequential Monte-Carlo scheme, or ABC-SMC~\citep{Sisson2009}, implemented in the \texttt{pyABC}\footnote{https://pyabc.readthedocs.io} Python 
package~\citep{Klinger2018}.
Briefly, this approach uses numerical stochastic simulations of the model to infer a posterior distribution over the model parameters. It is a method of likelihood-free, simulation-based inference~\citep{Cranmer2020}, that is, for estimating a posterior distribution when a likelihood function cannot be computed. It is therefore suitable to our case, in which the likelihood function can only be approximated from simulations, and cannot be directly computed. 
%adapted/expanded description from Syga et al. 2021:
In short, the algorithm employs sequential importance sampling over generations. 
In generation $t$, the algorithm draws sets of parameters $\theta_i$ from a given proposal distribution and consequently simulates data $C_i$ from the model until $n_t$ instances were accepted based on the comparison to observed data $C_{obs}$ via a distance function $D(C_i, C_{obs})$ and acceptance threshold $\epsilon_t$, such that $D(C_i, C_{objs})\leq{\epsilon_t}$. As the distance function we choose $1-\likelihood$. For the acceptance threshold the median of the distances of the accepted instances of the previous generation are applied (for the first generation the acceptance threshold is defined $0.01$). 
When $n_t$ instances are accepted new proposal distribution is constructed by multivariate normal kernel with adaptive covariance matrix. In the first generation the proposal distribution equals to the prior. 
The size of $n_t$ is determined by adaptive population strategy~\citep{Klinger2017}.
We stop the iterations, when no more sufficient change in acceptance threshold occur, and no change in posterior.
After receiving weighted particles of the posterior distribution using ABC-SMC scheme, we calculate Kernel Density Estimate using Gaussian kernels. $50,000$ sampled points from KDE represent the posterior distribution over the model parameters, and we use it for plots, median calculation, etc.
% TODO convergence tests, etc. Number of replicates for different models.

@@

We used growth curves of $2n+1$ and $2n$ in \SI{39}{\celsius} to estimate  $w_{2n+1}$. We used the same prior for $w_{2n+1^*}$ and $w_{2n^*}$, assuming that fitness values shouldn't differ too much. We also tested different priors, but this one (let's call it main prior) was better in the sense of WAIC, prediction plots, estimated parameters. %TODO provide formula of WAIC. Table? To show ppc plots also?. 
\\We tested uniform prior (let's call it uninformative), $\mathit{Uniform}(1,6)$, for all $w_{2n+1}$, $w_{2n+1^*}$, $w_{2n^*}$ and only for $w_{2n+1^*}$, $w_{2n^*}$, leaving $w_{2n+1}$ prior as before. In both scenarios the estimates by the model parameters of fitness looks too high: for the first scenario the median estimates were  $w_{2n+1}=2.767$, $w_{2n+1^*}=5.79$, $w_{2n^*}=5.827$ and for the second $w_{2n+1}=1.046$, $w_{2n+1^*}=4.642$, $w_{2n^*}=4.684$. We don't expect to have such high fitness values. Moreover the WAIC for the models with these priors are 0.57 and 1.9 respectively, while WAIC for the main prior model is 0.27. Also for the model with uninformative prior for all $w_{2n+1}$, $w_{2n+1^*}$, $w_{2n^*}$ the fixation time of 2n+1 is about 16 in average, which contradicts unpublished observations that were about 250-300.
\\Also we tried to use additional growth curves. We used growth curves of 2n* (\emph{refined} from \citet{Yona2012}) and 2n+1 in \SI{39}{\celsius} to estimate $w_{2n*}/w_{2n+1}$. The same prior was used for $w_{2n*}/w_{2n+1*}$. This model didn't perform well, it resulted in WAIC 2.8.
\\Another try was to use growth curves of $2n+1$ and $2n$ in \SI{30}{\celsius} to estimate $1-c$. This estimation assumes that the cost of aneuploidy is the same in \SI{39}{\celsius} and \SI{30}{\celsius}; this might be incorrect, but we only assumed this to generate a prior distribution for the fitness values. The prior for $b$ was taken the same as for $c$. This model didn't work at all: no parameter sets were found with the likelihood greater than zero.
%TODO it make sense because bla bla?

Because \texttt{Curveball} uses a maximum-likelihood approach to estimate model parameters, we were able to estimate a distribution of relative fitness values by sampling from a truncated multivariate normal distribution, defined by the maximum-likelihood covariance matrix. Thus, we sampled 10,000 values for appropriate fitness, which we used as prior distributions. 
See Figures~\ref{fig:growth-curves-30deg} and~\ref{fig:growth-curves-39deg}.

For the mutation rate $\mu$ and aneuploidy rate $\delta$ we used uniform priors, $\mathit{Uniform}(1e-8,1e-5)$ and $\mathit{Uniform}(1e-6,1e-3)$ respectively. We tried also uniform priors with lower than the $1e-8$ bound for $\mu$, but anyway the posterior ended with higher values. 
For the single locus model without aneuploidy, wider uniform prior for the mutation rate was used - $\mathit{Uniform}(1e-10,1e-5)$. 
%TODO ILIA please report here the point estimate and 95% CI for the three distributions, I only have for 1-c+b: 0.9863 (95% CI 0.9790 - 0.9939).
\paragraph{Model comparison.}  % TODO ILIA what did you do for model comparison/selection? WAIC

% Results
\section*{Results}
